Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2023-06-19 unimodular-covers-CombTheory-final.tex
solved the last issue

%
Since we can perform without loss of generality respective lattice translations to $P$ and to $Q$, we assume that the segments $p$ and $q$ intersect in their relative interiors and that $f_q(p_1)=-1$. That is,
\[
-1=f_q(p_1) < 0 < f_q(p_1), \quad
-f_p(q_1) < 0 < f_p(q_2).
\]
In particular, $p$ must be in the interior of $q +\vecline p$ and $q$ in the interior of $p +\vecline q$
Our assumption also implies that $p_1$ is the unique lattice point with $f_q(x)=-1$ in $q +\vecline p$. Similarly, the unique lattice point in the strip with $f_q(x)=1$ is  $q_1+q_2 -p_1$.

@@

Let $g$ be the primitive lattice functional constant on $[p_1, q_2]$ (and therefore constant also on $[q_1, q_1+q_2-p_1]$). By the assumption that $f_q(p_1)=-1$ the quadrilateral spanned by these two parallel segments is unimodular, so  the values of $g$ on them differ by $1$. We choose the sign of $g$ so that 
\[
g([p_1, q_2])= g( [q_1, q_1+q_2-p_1]) -1. 
\]

@@

\begin{proof}
Since $b$ and $t$ must respectively separate $p_1$ and $q_1+q_2-p_1$ from the other two vertices of the parallelogram $\conv(q_1, p_1, q_2, q_1+q_2-p_1)$, they must respectively intersect its (parallel) edges $[p_1, q_2]$ and $[q_1, q_1+q_2-p_1]$, which implies the stated inequalities for $b$ and $t$.
The same argument  applied to the parallelogram  $\conv(p_1, q_2, p_2, p_1+p_2-q_2)$, yields the inequalities for $r$.
\end{proof}

\begin{figure}[htb]
\scalebox{.75}{\input{claim4.pdf_t}}
\caption{Illustration of the proof of \Cref{claim:g}}
\label{fig:claim4}
\end{figure}

\begin{claim}
\label{claim:g2}
$g(r_2) - g(r_1) <1$.
\end{claim}

\begin{proof}
UNFINISHED

Since the normal fan of $Q$ refines that of $P$, $Q$ must contain an edge parallel to $r$ on the right. Let $r'$ be a primitive segment along this edge, so that $r'$ is a translated copy of $r$.
 Let $r_1'$ and $r_2'$ be its endpoints.

Consider the segment $s$ contained in $r_1' + \vecline p$ with endpoints $s_1=r_1'$ and $s_2$ on the line spanned by $t$. The width of $s$ with respect to $g$ is $g(s_2)-g(s_1) <1$, because $s$ is shorter than $Q \cap p$, which is strictly contained between the consecutive parallel lattice lines through $p_1$ and $q_2$ and through $q_1$ and $q_1+q_2-p_1$ where $g$ is constant.

We now observe that $g(r'_2) < g(s_2)$, because $r'_2$ lies to the right of $s_2$ ($r$ has positive slope) and below the line spanned by $t$, and $g$ decreases moving to the right along $t$ (\Cref{claim:g}). Thus $g(r_2)-g(r_1) = g(r'_2)-g(r'_1) < g(s_2)-g(s_1) <1$, that is, $g(r_2)=g(r_1)$. This contradicts \Cref{claim:g}.
%Since $t$ separates $q_1$ and $q_1+q_2-p_1$ and $g$ decreases from $t_1$ to $t_2$ (by \Cref{claim:g}), the inequality $g(x)< g(d')$ holds on $Q\cap V_2$, and in particular for $r_2'$. Since $r_2'$ is a lattice point, $g(r_2')\leq g(d)= g(r_1')$, which contradicts \Cref{claim:g}).
\end{proof}

We now have a contradiction. Claim~\ref{claim:g} says that $g(r_2) - g(r_1) >0$ and Claim~\ref{claim:g2}  that $g(r_2) - g(r_1) <1$. Since  $g(r_2) - g(r_1)$ is an integer, this is impossible.
\end{proof}

>>>>

%
Since we can perform without loss of generality respective lattice translations to $P$ and to $Q$, we assume that the segments $p$ and $q$ intersect in their relative interiors and that $f_q(p_1)=-1$. That is,
\[
-1=f_q(p_1) < 0 < f_q(p_1), \quad
f_p(q_1) < 0 < f_p(q_2).
\]
In particular, $p$ must be in the interior of $q +\vecline p$ and $q$ in the interior of $p +\vecline q$
Our assumption also implies that $p_1$ is the unique lattice point with $f_q(x)=-1$ in $q +\vecline p$. Similarly, the unique lattice point in the strip with $f_q(x)=1$ is  $q_1+q_2 -p_1$.

@@

Let $g$ be the primitive lattice functional constant on $[p_1, q_2]$ (and therefore constant also on $[q_1, q_1+q_2-p_1]$). By the assumption that $f_q(p_1)=-1$ the quadrilateral $:=\conv(q_1, p_1, q_2, q_1+q_2-p_1)$
 is unimodular, so  the values of $g$ on the edges $[p_1, q_2]$ and $[q_1, q_1+q_2-p_1]$ differ by $1$. We choose the sign of $g$ so that 
\[
g([p_1, q_2])= g( [q_1, q_1+q_2-p_1]) -1. 
\]

@@

\begin{proof}
Since $b$ and $t$ must respectively separate $p_1$ and $q_1+q_2-p_1$ from the other two vertices of the parallelogram $\conv(q_1, p_1, q_2, q_1+q_2-p_1)$, they must respectively intersect its (parallel) edges $[p_1, q_2]$ and $[q_1, q_1+q_2-p_1]$, which implies the stated inequalities for $b$ and $t$.
The same argument  applied to the parallelogram  $\conv(p_1, q_2, p_2, p_1+p_2-q_2)$, yields the inequality for $r$.
\end{proof}

\begin{figure}[htb]
\scalebox{.75}{\input{claim4.pdf_t}}
\caption{Illustration of the proof of \Cref{claim:g}. The functional $g$ is constant along the dashed diagonal lines}
\label{fig:claim4}
\end{figure}

We are now ready to show a contradiction. Since the normal fan of $Q$ refines that of $P$, $Q$ must contain an edge parallel to $r$ on the right. Let $r'$ be a primitive segment along this edge, so that $r'$ is a translated copy of $r$.
 Let $r_1'$ and $r_2'$ be its endpoints.

We consider two cases, depending on whether $q_1+q_2-p_1$ lies ``to the right'' or ``to the left'' of $p_1$. That is, whether $f_p(q_1+q_2-p_1)$ is positive or negative. %For illustration use the left part of Figure~\ref{fig:claim4} in which $f_p(q_1+q_2-p_1)=0$, and in which both arguments work:

\begin{itemize}
\item \underline{$f_p(q_1+q_2-p_1)> 0$}. In this case the segment $r$ in the boundary of $P$ separates $p$ from  $q_1+q_2-p_1$.
Let $r''_1$ and $r''_2$ be the intersection points of the line containing $r$ with $b$ and $t$ respectively. We have that
$r'=[r'_1,r'_2]$ is parallel to $[r''_1,r''_2]$ but shorter, so that
\[
w= \width_{f_q}(p) \le \width_{f_q}(r) = \width_{f_q}(r') \le  f_q(r''_2) -f_q(r''_1), 
\]
where the first inequality comes from \eqref{eq:widthp}.
But on the other hand
\[
f_q(r''_2) < f_q(q_1+q_2-p_1) = 1, \quad f_q(r''_1) > f_q(p_1) =-1,
\]
which leads to the contradiction $w<2$.

\item \underline{$f_p(q_1+q_2-p_1)\ge 0$}.
Consider the segment $s$ contained in $r_1' + \vecline p$ with endpoints $s_1=r_1'$ and $s_2$ on the line spanned by $t$. The width of $s$ with respect to $g$ is $g(s_2)-g(s_1) <1$, because $s$ is shorter than $Q \cap p$, which is strictly contained between the consecutive parallel lattice lines through $p_1$ and $q_2$ and through $q_1$ and $q_1+q_2-p_1$ where $g$ is constant.

We now observe that $g(r'_2) < g(s_2)$, because $r'_2$ lies to the right of $s_2$ ($r$ has positive slope) and below the line spanned by $t$, and $g$ decreases moving to the right along $t$ (\Cref{claim:g}). Thus $g(r_2)-g(r_1) = g(r'_2)-g(r'_1) < g(s_2)-g(s_1) <1$. Since $g(r_2)$ and $g(r_1)$ are integers this leads to $g(r_2)\le g(r_1)$, in contradiction with \Cref{claim:g}.
%Since $t$ separates $q_1$ and $q_1+q_2-p_1$ and $g$ decreases from $t_1$ to $t_2$ (by \Cref{claim:g}), the inequality $g(x)< g(d')$ holds on $Q\cap V_2$, and in particular for $r_2'$. Since $r_2'$ is a lattice point, $g(r_2')\leq g(d)= g(r_1')$, which contradicts \Cref{claim:g}).
\end{itemize}
\end{proof}
